EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-08-20 jlamp/jlamp.tex
minima

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label{rem:length} It is well known that $(\mathbb{N}), \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathbb{N}\to 1^\star$. In particular, for every set $X$ the unique arrow $!_\X\colon X\to 1$ into the terminal objects induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside. 
		\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=8pt@C=30pt{1 \ar[r]^{\id{1}} \ar@{>->}[d]^{\delta_n}& 1 \ar@{>->}[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=8pt@C=28pt{\mathbb{N} \ar[r]^{\Delta} \ar@{<.}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

@@

\noindent
\begin{minipage}[r]{.83\linewidth}
	\begin{rem}\label{rem:len} It is worth noticing that for $X$ a set and every $n\in \mathbb{N}$, $X^n$ is the preimage of $n$ along $\lgh_{X}$. This amount to say that the square aside is a pullback. This follows at once from the fact that $\Set$ is \emph{extensive} (see \cite{carboni1993introduction} and also \Cref{word:ext}). 
	\end{rem}
\end{minipage}\hfill \begin{minipage}[l]{.15\linewidth} \vspace{-.2cm}
\xymatrix@R=13pt{X^n \ar@{>->}[r]^{v_n} \ar[d]^{!_{X^n}}& X^\star  \ar[d]^{\lgh_{X}}\\ 1 \ar@{>->}[r]^{\delta_n} & \mathbb{N}}
\end{minipage}

@@

\begin{rem}\label{rem:con}Notice that  $\lmo$ is not complete: it lacks e.g. a terminal object. To see this, notice that the $\mathcal{H}:=(\emptyset, 1, ?_{1^\star}, 1_{1^\star} )$ is left-monogamous and that 
	$\lmo(\mathcal{H}, \mathcal{G})$ is in bijection with $V_{\mathcal{G}}$ for every hypergraph $\mathcal{G}$. Thus if  $\mathcal{T}$ is terminal in $\lmo$, then its set of vertices $V_{\mathcal{T}}$ must have cardinality $1$.  We have two possibilities

@@

\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.
We can further take the coproduct $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}})$ of the family $\{V_{e}\}_{e\in E_\mathcal{G}}$ and define
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{n_e + m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of $1$. 
\marginpar{ovvero?}

@@

	We can then depict $\mathcal{G}^\Sigma $ and $\sp(\mathcal{G}^\Sigma )$ as below\marginpar{e le lettere?}
	\begin{center}	\begin{tikzpicture}
			\begin{pgfonlayer}{nodelayer}
				\node[style=node] (v) at (0, 0){};
				\node[style=none] (vlab) at (0, -0.4){$v$};
				\node[style=small box] (n) at (-1.5, 0) {};
				\node[style=none] (nlab) at (-1.5, 0.5) {$e$};
				\node[style=none] (nattach) at (-1.6, 0) {};

@@

\noindent 
\begin{minipage}[l]{.68\linewidth}
\begin{rem}Notice that $\sp$ is neither right nor left adjoint to $S$.
	\begin{itemize}
		\item To see that $\sp$ cannot be a right adjoint to $S$, consider the ``double loop'' $\mathcal{D}=(2, 1, !_2, !_2)$ depicted aside. $\sp(\mathcal{D})$ is then isomorphic to the graph $\mathcal{G}=(2, 4, s_{\mathcal{G}}, t_{\mathcal{G}})$ with
		\[s_{\mathcal{G}}(0) = 0 \quad  t_{\mathcal{G}}(0) =  1 \quad  s_{\mathcal{G}}(1) = 2 \quad  t_{\mathcal{G}}(1) = 3 \]

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_1, ?_1)$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular, this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-1, 1) {};
			\node[style=none] (timeslab) at (-1, 1.5) {$0$};
			\node[style=none] (timesin) at (-1, .7) {};
			\node[style=none] (timesout) at (-.7, 1){};
			\node[style=small box] (div) at (1, 1) {};
			\node[style=none] (divlab) at (1, 1.5) {$1$};
			\node[style=none] (divin) at (1, .7) {};
			\node[style=none] (divout) at (.7, 1){};

@@

\begin{exa}\label{ex:noterm}
In general $\shyp$ is not complete. For instance it does not have a terminal object. Too see this notice that the $\mathcal{H}:=(\emptyset, 1, ?_{1^\star}, 1_{1^\star} )$ is separated and that 
$\shyp(\mathcal{H}, \mathcal{G})$ is in bijection with $V_{\mathcal{G}}$ for every hypergraph $\mathcal{G}$. Thus if  $\mathcal{T}$ is terminal in $\shyp$, then its set of vertices $V_{\mathcal{T}}$ must have cardinality $1$.

@@

In turn, by \Cref{thm:slice-functors}, Corollaries \ref{cor:eq}, \ref{cor:pusho} and \Cref{lem:regmon} we get the following result

\begin{thm} Let $\mathcal{G}$ be an hypergraph. Then the following fact hold
	\begin{enumerate}
		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		\item $\lmo_{\mathcal{G}}$ is quasiadhesive.
	\end{enumerate}
\end{thm}

@@

\newpage
	\smallskip\noindent 
	\parbox{4cm}{ \xymatrix{A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}}\hfill \parbox{8.5cm}{$(\Leftarrow)$ Since $h$ and $k$ are monos then we have the two pullback squares on the left. Thus by \Cref{lim} the pullback of $(h,k)$ along itself has isomorphisms as projections and so $(h,k)$ is mono. \qedhere}

@@

\newpage
\begin{proof} 
Let $*$ be the only object of the category $\T$ and define functors $F_1\colon \comma{\id{X}}{\delta_X}\to \X/X$ and $G_1\colon \X/X\to \comma{\id{X}}{\delta_X}$ as follows
	\[	\begin{split}
		\functor[l]{(Y, *,  f)}{(h, \id{*} )}{( Y', *,  g)}
		\functormapsto
		\rfunctor{f }{h}{g}
	\end{split} \qquad \begin{split}
		\functor[l]{f }{h}{g}
		\functormapsto
		\rfunctor{(\dom(f), *, f)}{(h, \id{*} )}{(\dom(g), *, g)}
	\end{split}\]
	Similarly, we have $F_2\colon \comma{\delta_X}{\id{X}}\to X/\X$ and $G_2\colon X/\X\to \comma{\delta_X}{\id{X}}$
	\[	\begin{split}
		\functor[l]{(*, Y, f)}{(\id{*},h )}{(*, Y', g)}
		\functormapsto
		\rfunctor{f }{h}{g}
	\end{split} \qquad \begin{split}
		\functor[l]{f }{h}{g}
		\functormapsto
		\rfunctor{(*, \cod(f), f)}{(\id{*}, h )}{(*, \cod(g), g)}
	\end{split}\]
	It is now obvious to see that $F_1,G_1$ and $F_2, G_2$ are pairs of inverses.   
\end{proof}

@@

In particular, if $\X$ has pullbacks, equalizers or pushouts, then for every object $X$ the slice $\X/X$ has such limits and colimits.

\iffalse
\todo{Tutta la parte qua sotto non mi è chiaro a cosa serva}
In a category $\cat X$ with pullbacks, each $f: X \to Y$ induces a functor $f^*: \cat X/Y \to \cat X/X$,
which sends each morphism $a: A \to Y$ of $\cat X$ onto its pullback along $f$, $p_a$, and each morphism $h: a \to b$ onto the unique arrow from the pullback of $a$ along $f$ to the pullback of $b$ along $f$.

>>>>

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label{rem:length} It is well known that $(\mathbb{N}, \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathbb{N}\to 1^\star$. In particular, for every set $X$ the unique arrow $!_\X\colon X\to 1$ into the terminal object induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside. 
		\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=8pt@C=30pt{1 \ar[r]^{\id{1}} \ar@{>->}[d]^{\delta_n}& 1 \ar@{>->}[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=8pt@C=28pt{\mathbb{N} \ar[r]^{\Delta} \ar@{<.}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

@@

\noindent
\begin{minipage}[r]{.83\linewidth}
	\begin{rem}\label{rem:len} It is worth noticing that for $X$ a set and every $n\in \mathbb{N}$, $X^n$ is the preimage of $n$ along $\lgh_{X}$. This amounts to say that the square aside is a pullback. This follows at once from the fact that $\Set$ is \emph{extensive} (see \cite{carboni1993introduction} and also \Cref{word:ext}). 
	\end{rem}
\end{minipage}\hfill \begin{minipage}[l]{.15\linewidth} \vspace{-.2cm}
\xymatrix@R=13pt{X^n \ar@{>->}[r]^{v_n} \ar[d]^{!_{X^n}}& X^\star  \ar[d]^{\lgh_{X}}\\ 1 \ar@{>->}[r]^{\delta_n} & \mathbb{N}}
\end{minipage}

@@

\begin{rem}\label{rem:con}Notice that  $\lmo$ is not complete: it lacks e.g. a terminal object. To see this, notice that the $\mathcal{H}:=(\emptyset, 1, ?_{1}, ?_{1} )$ is left-monogamous and that 
	$\lmo(\mathcal{H}, \mathcal{G})$ is in bijection with $V_{\mathcal{G}}$ for every hypergraph $\mathcal{G}$. Thus if  $\mathcal{T}$ is terminal in $\lmo$, then its set of vertices $V_{\mathcal{T}}$ must have cardinality $1$.  We have two possibilities

@@

\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.
We can further take the coproduct $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}})$ of the family $\{V_{e}\}_{e\in E_\mathcal{G}}$ and define
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{n_e + m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of the terminal object $1$. 
\marginpar{ovvero?}

@@

	We can then depict $\mathcal{G}^\Sigma $ and $\sp(\mathcal{G}^\Sigma )$ as below
	\begin{center}	\begin{tikzpicture}
			\begin{pgfonlayer}{nodelayer}
				\node[style=node] (v) at (0, 0){};
				\node[style=none] (vlab) at (0, -0.4){$v$};
				\node[style=small box] (n) at (-1.5, 0) {};
				\node[style=none] (nlab) at (-1.5, 0.5) {$e$};
				\node[style=none] (nattach) at (-1.6, 0) {};

@@

\noindent 
\begin{minipage}[l]{.68\linewidth}
\begin{rem}Notice that $\sp$ is neither right nor left adjoint to $S$.
	\begin{itemize}
		\item To see that $\sp$ cannot be a right adjoint to $S$, consider the ``double loop'' $\mathcal{D}=(2, 1, !_2, !_2)$ depicted aside. Then $\sp(\mathcal{D})$ is isomorphic to the graph $\mathcal{G}=(2, 4, s_{\mathcal{G}}, t_{\mathcal{G}})$ with
		\[s_{\mathcal{G}}(0) = 0 \quad  t_{\mathcal{G}}(0) =  1 \quad  s_{\mathcal{G}}(1) = 2 \quad  t_{\mathcal{G}}(1) = 3 \]

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_{1^\star}, ?_{1^\star})$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular, this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-1, 1) {};
			\node[style=none] (timeslab) at (-1, 1.5) {$0$};
			\node[style=none] (timesin) at (-1, .7) {};
			\node[style=none] (timesout) at (-.7, 1){};
			\node[style=small box] (div) at (1, 1) {};
			\node[style=none] (divlab) at (1, 1.5) {$1$};
			\node[style=none] (divin) at (1, .7) {};
			\node[style=none] (divout) at (.7, 1){};

@@

\begin{exa}\label{ex:noterm}
In general $\shyp$ is not complete. For instance it does not have a terminal object. Too see this notice that the $\mathcal{H}:=(\emptyset, 1, ?_{1^\star}, ?_{1^\star} )$ is separated and that 
$\shyp(\mathcal{H}, \mathcal{G})$ is in bijection with $V_{\mathcal{G}}$ for every hypergraph $\mathcal{G}$. Thus if  $\mathcal{T}$ is terminal in $\shyp$, then its set of vertices $V_{\mathcal{T}}$ must have cardinality $1$.

@@

In turn, by \Cref{thm:slice-functors}, Corollaries \ref{cor:eq}, \ref{cor:pusho} and \Cref{lem:regmon} we get the following result.

\begin{thm} Let $\mathcal{G}$ be an hypergraph. Then %the following fact hold
%	\begin{enumerate}
%		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		%\item 
		$\lmo_{\mathcal{G}}$ is quasiadhesive.
%	\end{enumerate}
\end{thm}

@@

%\newpage
	\smallskip\noindent 
	\parbox{4cm}{ \xymatrix{A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}}\hfill \parbox{8.5cm}{$(\Leftarrow)$ Since $h$ and $k$ are monos then we have the two pullback squares on the left. Thus by \Cref{lim} the pullback of $(h,k)$ along itself has isomorphisms as projections and so $(h,k)$ is mono. \qedhere}

@@

\begin{proof} 
Let $1$ be the only object of the category $\T$ and define functors $F_1\colon \comma{\id{X}}{\delta_X}\to \X/X$ and $G_1\colon \X/X\to \comma{\id{X}}{\delta_X}$ as follows
	\[	\begin{split}
		\functor[l]{(Y, 1,  f)}{(h, \id{1} )}{( Y', 1,  g)}
		\functormapsto
		\rfunctor{f }{h}{g}
	\end{split} \qquad \begin{split}
		\functor[l]{f }{h}{g}
		\functormapsto
		\rfunctor{(\dom(f), 1, f)}{(h, \id{1} )}{(\dom(g), 1, g)}
	\end{split}\]
	Similarly, we have $F_2\colon \comma{\delta_X}{\id{X}}\to X/\X$ and $G_2\colon X/\X\to \comma{\delta_X}{\id{X}}$
	\[	\begin{split}
		\functor[l]{(1, Y, f)}{(\id{1},h )}{(1, Y', g)}
		\functormapsto
		\rfunctor{f }{h}{g}
	\end{split} \qquad \begin{split}
		\functor[l]{f }{h}{g}
		\functormapsto
		\rfunctor{(1, \cod(f), f)}{(\id{1}, h )}{(1, \cod(g), g)}
	\end{split}\]
	It is now obvious to see that $F_1,G_1$ and $F_2, G_2$ are pairs of inverses.   
\end{proof}

@@

%In particular, if $\X$ has pullbacks, equalizers or pushouts, then for every object $X$ the slice $\X/X$ has such limits and colimits.

\iffalse
\todo{Tutta la parte qua sotto non mi è chiaro a cosa serva}
In a category $\cat X$ with pullbacks, each $f: X \to Y$ induces a functor $f^\*: \cat X/Y \to \cat X/X$,
which sends each morphism $a: A \to Y$ of $\cat X$ onto its pullback along $f$, $p_a$, and each morphism $h: a \to b$ onto the unique arrow from the pullback of $a$ along $f$ to the pullback of $b$ along $f$.
